\mytitle

### 2016-11-02 pdocs/thesis/article-conclusion.tex
bullet def changed to newcommand cbullet

There is also work to be done in the area of performance optimization.
One possible area is the reuse of once calculated values if memory capacity is not limited.
This could done, e.g., when dealing with high frequency modes.
Another area is usage of the graphics processing unit (GPU) for calculation.
\bullet\ already has an experimental GPU pipeline but most constraint types are not 
yet supported. GPU support was one reason for selecting
\bullet\ for use in this study

Integration to 3D modelling and animation software products 
should also be solved before we can expect to
see realistic plasticity in main stream games. 
\bullet\ is already integrated into several ones like Blender 
and it was another significant reason for selecting
\bullet\  to be used in this study.

>>>>

There is also work to be done in the area of performance optimization.
One possible area is the reuse of once calculated values if memory capacity is not limited.
This could done, e.g., when dealing with high frequency modes.
Another area is usage of the graphics processing unit (GPU) for calculation.
\cbullet\ already has an experimental GPU pipeline but most constraint types are not 
yet supported. GPU support was one reason for selecting
\cbullet\ for use in this study

Integration to 3D modelling and animation software products 
should also be solved before we can expect to
see realistic plasticity in main stream games. 
\cbullet\ is already integrated into several ones like Blender 
and it was another significant reason for selecting
\cbullet\  to be used in this study.

### 2016-11-02 pdocs/thesis/article-introduction.tex
acm tog styling of article started
acm tog styling of article continues
bullet def changed to newcommand cbullet

% !TeX root = article.tex
\section{Introduction}

Computational techniques used in modern computer games are able to describe
complex dynamic systems such as those found in military and other motorized vehicles
with a high level of accuracy while still being able to solve the equations of motion in real time.
Film production and the computer game genre of war games
are two areas that have greatly benefited from rapidly improving simulation technology. 
In practice, software for such games is often written using open-source physics engine platforms  
like ODE - \cite{ode}, \bullet\ - \cite{bullet} and Box2D - \cite{box2d}.

@@

This study will introduce an approach to account for plastic deformation in game applications.   
In the introduced method, plastic deformation takes place if the force or moment exceeds a predefined 
limit, deformation absorbs energy and joint breaks if plastic capacity is exceeded. 
The approach is based on using joint motors to model plasticity. 
The study extends a method introduced by
\citet{erleben.thesis} 
%\citet[p.~90]{erleben.thesis} 
which was originally proposed for modelling friction in joints. 
In the introduced method adjacent bodies are connected by motors. 
Motor power production limits are estimated based on the plastic section modulus. 
Joint breaking is based on summing plastic deformation and comparing it to a
predefined material based limit. The elastic part of deformation is modelled by employing 
a spring based on modification of an existing constraint in \bullet.

>>>>

% !TeX root = article.tex
\begin{abstract}
We introduce efficient method to simulate ductile fracture in existing physics engines.
\end{abstract}

\section{Introduction}

Computational techniques used in modern computer games are able to describe
complex dynamic systems such as those found in military and other motorized vehicles
with a high level of accuracy while still being able to solve the equations of motion in real time.
Film production and the computer game genre of war games
are two areas that have greatly benefited from rapidly improving simulation technology. 
In practice, software for such games is often written using open-source physics engine platforms  
like ODE - \cite{ode}, \cbullet\ - \cite{bullet} and Box2D - \cite{box2d}.

@@

This study will introduce an approach to account for plastic deformation in game applications.   
In the introduced method, plastic deformation takes place if the force or moment exceeds a predefined 
limit, deformation absorbs energy and joint breaks if plastic capacity is exceeded. 
The approach is based on using joint motors to model plasticity. 
The study extends a method introduced by
\cite{erleben.thesis} 
%\citet[p.~90]{erleben.thesis} 
which was originally proposed for modelling friction in joints. 
In the introduced method adjacent bodies are connected by motors. 
Motor power production limits are estimated based on the plastic section modulus. 
Joint breaking is based on summing plastic deformation and comparing it to a
predefined material based limit. The elastic part of deformation is modelled by employing 
a spring based on modification of an existing constraint in \cbullet.

### 2016-11-02 pdocs/thesis/article-section-2.tex
bullet def changed to newcommand cbullet

Work presented by  
\cite{erleben.thesis} provides the basis for the velocity-based formulation discussed in this work.
%\citet[p.~45-50]{erleben.thesis}.
% pdf page 64
In the following section, these formulations will be clarified by a simple example using \bullet\ implementation.

@@

Constraint processing in \bullet\ is based on ODE, \cite{ode}.
Joints are also discussed in detail in  
\cite{erleben.thesis}.
%\citet[p.~60-90]{erleben.thesis}.
Equations \ref{eq:constraintEquation}, \ref{eq:lambdaLow} and
\ref{eq:lambdaHigh} 
are created for each constraint. 
Derivation for terms in Equation \ref{eq:constraintEquation}
can be done using the position and orientation of connected bodies
e.g. for ball joint formulation is based on both joint points having the same position.
In contact cases, formulation is easier if it is done using velocities, \cite{ode.joints}.

@@

In the following section, these equations will be explained by a simple example.
The main parameters  and corresponding fields in \bullet\  
are given in Table \ref{tab:constraintParameters}.

@@

Further enhancements were created and tested by forking \bullet\ source code
and adding new constraints, \cite{pbullet}.
Instructions for using  windows executable and  source code are available, \cite{bp}.

>>>>

Work presented by  
\cite{erleben.thesis} provides the basis for the velocity-based formulation discussed in this work.
%\citet[p.~45-50]{erleben.thesis}.
% pdf page 64
In the following section, these formulations will be clarified by a simple example using \cbullet\ implementation.

@@

Constraint processing in \cbullet\ is based on ODE, \cite{ode}.
Joints are also discussed in detail in  
\cite{erleben.thesis}.
%\citet[p.~60-90]{erleben.thesis}.
Equations \ref{eq:constraintEquation}, \ref{eq:lambdaLow} and
\ref{eq:lambdaHigh} 
are created for each constraint. 
Derivation for terms in Equation \ref{eq:constraintEquation}
can be done using the position and orientation of connected bodies
e.g. for ball joint formulation is based on both joint points having the same position.
In contact cases, formulation is easier if it is done using velocities, \cite{ode.joints}.

@@

In the following section, these equations will be explained by a simple example.
The main parameters  and corresponding fields in \cbullet\  
are given in Table \ref{tab:constraintParameters}.

@@

Further enhancements were created and tested by forking \cbullet\ source code
and adding new constraints, \cite{pbullet}.
Instructions for using  windows executable and  source code are available, \cite{bp}.

### 2016-11-02 pdocs/thesis/article-section-3.tex
bullet def changed to newcommand cbullet

Method btSequentialImpulseConstraintSolver::solveGroupCacheFriendlySetup
in \bullet\ is used to get values for the internal variables described in Figure \ref{fig:solVars}.

\begin{figure}
\begin{description}
\item[velError] is calculated using velocities and external impulses of the connected bodies.  
 \begin{description}
\item[In constraint cases,] the main contibutor is the relative speed of the bodies at the joint point.
\item[In contact cases,] the main contributor is the relative speed of the bodies at the point of contact. 
Bodies are not allowed to penetrate each other. Restitution increases velError.
If contact is not penetrating velError is reduced by $penetration * timeStep$.
\end{description}
\item[posError] is provided by the constraint. It is a significant factor in designing stable constraints.
 \begin{description}
 \item[In fixed cases,] the value is about 12 times the actual position error. Factor 12 is based on the time step (60) 
 and the default value of error reduction parameter (erp), which has a value of 0.2 in this context.
 \item[In elastic plastic cases,]  the value is set to zero if the impulse would be larger the than maximum impulse or
the spring simulation cannot be done in a stable way.
 \item[In contact cases,] the value is zero if there is no penetration. For penetration cases it is 
$\frac{-penetration\, erp}{timeStep}$. In contact cases the default value for erp is 0.8.
 \end{description}
\item[rhs (c)] is calculated by velError jInv + posError jInv
\item[jInv] is calculated using masses and inertias of connected bodies and constraint geometry. 
 \begin{description}
\item[In constraint cases,] it is the mass of the body (6000).
\item[In contact cases,] it varies below the mass of the body.
\end{description}
\item[Impulse] is impulse applied to the body during a timestep.
 \begin{description}
\item[In constraint cases,] it is obtained from the btJointFeedback structure.
\item[In contact cases,] it is obtained by summing the applied impulses from the active manifolds.
\end{description}
\item[erp] Error reduction parameter (0...1) is used to handle numerical issues e.q. body drifting. 
Setting erp to 1 would in theory eliminate error in one step but in practice a value of 0.2 - 0.8 is used in most cases.
\end{description}
\caption{Internal variables used in \bullet\ in constraint solving.}
\label{fig:solVars}
\end{figure}

@@

There are currently two alternative six-dof-spring constraint implementations in \bullet\ and 
in this work elastic-plastic versions of both of them are developed.

@@

\bullet\ simulations are usually done using a fixed time step of 1/60 s, i.e., 16.67 ms. 
For this case that is too large. 
The default time step was selected to be 5 ms outside impact  and 0.1 ms during impact. 
An automatic time stepping routine changes the timestep so that at angles higher than 0.2 radians 5 ms time step is
used and adjusts it linearly to a selected time step between the angles of 0.2 and 0.05 radians.

>>>>

Method btSequentialImpulseConstraintSolver::solveGroupCacheFriendlySetup
in \cbullet\ is used to get values for the internal variables described in Figure \ref{fig:solVars}.

\begin{figure}
\begin{description}
\item[velError] is calculated using velocities and external impulses of the connected bodies.  
 \begin{description}
\item[In constraint cases,] the main contibutor is the relative speed of the bodies at the joint point.
\item[In contact cases,] the main contributor is the relative speed of the bodies at the point of contact. 
Bodies are not allowed to penetrate each other. Restitution increases velError.
If contact is not penetrating velError is reduced by $penetration * timeStep$.
\end{description}
\item[posError] is provided by the constraint. It is a significant factor in designing stable constraints.
 \begin{description}
 \item[In fixed cases,] the value is about 12 times the actual position error. Factor 12 is based on the time step (60) 
 and the default value of error reduction parameter (erp), which has a value of 0.2 in this context.
 \item[In elastic plastic cases,]  the value is set to zero if the impulse would be larger the than maximum impulse or
the spring simulation cannot be done in a stable way.
 \item[In contact cases,] the value is zero if there is no penetration. For penetration cases it is 
$\frac{-penetration\, erp}{timeStep}$. In contact cases the default value for erp is 0.8.
 \end{description}
\item[rhs (c)] is calculated by velError jInv + posError jInv
\item[jInv] is calculated using masses and inertias of connected bodies and constraint geometry. 
 \begin{description}
\item[In constraint cases,] it is the mass of the body (6000).
\item[In contact cases,] it varies below the mass of the body.
\end{description}
\item[Impulse] is impulse applied to the body during a timestep.
 \begin{description}
\item[In constraint cases,] it is obtained from the btJointFeedback structure.
\item[In contact cases,] it is obtained by summing the applied impulses from the active manifolds.
\end{description}
\item[erp] Error reduction parameter (0...1) is used to handle numerical issues e.q. body drifting. 
Setting erp to 1 would in theory eliminate error in one step but in practice a value of 0.2 - 0.8 is used in most cases.
\end{description}
\caption{Internal variables used in \cbullet\ in constraint solving.}
\label{fig:solVars}
\end{figure}

@@

There are currently two alternative six-dof-spring constraint implementations in \cbullet\ and 
in this work elastic-plastic versions of both of them are developed.

@@

\cbullet\ simulations are usually done using a fixed time step of 1/60 s, i.e., 16.67 ms. 
For this case that is too large. 
The default time step was selected to be 5 ms outside impact  and 0.1 ms during impact. 
An automatic time stepping routine changes the timestep so that at angles higher than 0.2 radians 5 ms time step is
used and adjusts it linearly to a selected time step between the angles of 0.2 and 0.05 radians.

### 2016-11-02 pdocs/thesis/bullet-plasticity.tex
bullet def changed to newcommand cbullet

\section{Simulation of Charpy impact test using \bullet}
\label{sec:bullet-charpy}

In following sections various methods for handling plasticity are described.
As physics engines have not been traditional area of research for plasticity, 
well known Charpy impact test was selected to be used as primary scenario
although it is against first two tips in \bullet Manual.

@@

\subsection{Introduction to constraint processing in \bullet}
Constraint processing in \bullet is based on ODE, \cite{ode}.
Mathematical background and detailed examples are available by \cite{ode.joints}.
Equations \ref{eq:constraintEquation}, \ref{eq:lambdaLow} and
\ref{eq:lambdaHigh} 
are created for each constraint.

@@

Main parameters  and corresponding fields in \bullet\  
 are described in table \ref{tab:constraintParameters}.

@@

\subsubsection{Breaking of constraints between bodies}
For standard \bullet constraints breakingImpulseThreshold (BITh) can be defined. If impulse is larger than set limit constraint is activated. 
This allows breaking to be simulated with fast calculation. 
This method is too simplified for ductile material and developing more precise and realistic methods is main objective of this work.

@@

\subsection{Spring2Constraint}
This basically same as SpringConstraint but uses newer btGeneric6DofSpring2Constraint which has recently appeared in
\bullet. 
Automatic stiffness limitation helps to avoid instability but may make constraint too soft.

@@

\section{Charpy impact test results using \bullet}
\label{sec:bullet-charpy-results}
Results are sensitive and they were obtained in February 2016 using following processes
\begin{itemize}
\item single precision debug build is used
\item sequential impulse constraint solver with 10 iterations is used
\item energy loss values are rounded to nearest 10 joules 
\end{itemize}

>>>>

\section{Simulation of Charpy impact test using \cbullet}
\label{sec:bullet-charpy}

In following sections various methods for handling plasticity are described.
As physics engines have not been traditional area of research for plasticity, 
well known Charpy impact test was selected to be used as primary scenario
although it is against first two tips in \cbullet Manual.

@@

\subsection{Introduction to constraint processing in \cbullet}
Constraint processing in \cbullet is based on ODE, \cite{ode}.
Mathematical background and detailed examples are available by \cite{ode.joints}.
Equations \ref{eq:constraintEquation}, \ref{eq:lambdaLow} and
\ref{eq:lambdaHigh} 
are created for each constraint.

@@

Main parameters  and corresponding fields in \cbullet\  
 are described in table \ref{tab:constraintParameters}.

@@

\subsubsection{Breaking of constraints between bodies}
For standard \cbullet constraints breakingImpulseThreshold (BITh) can be defined. If impulse is larger than set limit constraint is activated. 
This allows breaking to be simulated with fast calculation. 
This method is too simplified for ductile material and developing more precise and realistic methods is main objective of this work.

@@

\subsection{Spring2Constraint}
This basically same as SpringConstraint but uses newer btGeneric6DofSpring2Constraint which has recently appeared in
\cbullet. 
Automatic stiffness limitation helps to avoid instability but may make constraint too soft.

@@

\section{Charpy impact test results using \cbullet}
\label{sec:bullet-charpy-results}
Results are sensitive and they were obtained in February 2016 using following processes
\begin{itemize}
\item single precision debug build is used
\item sequential impulse constraint solver with 10 iterations is used
\item energy loss values are rounded to nearest 10 joules 
\end{itemize}

### 2016-11-02 pdocs/thesis/discussion.tex
Aki's comments on article and in general
bullet def changed to newcommand cbullet

\subsubsection{Implementation using GPU}
\bullet\ can already utilize GPU for many operations. 
Analysis on possible issues in used interfaces should be done before integrations.

@@

Current implementation is done based on \bullet\ but it still needs refactoring and testing before it can be 
accepted to main stream. Interfaces should be reafctored so that handling of additional features like large 
displacements and material nonlinearity can be implemented without further changes.

@@

Update of constraints can be done in  many ways but using callbacks allows clean integration i.e. no changes
to solver core are needed.
In \bullet\ there are already three per step callbacks.

@@

\subsection{Object subdivision just in time}
Just in time activation could avoid extra overhead from multiple objects and constraints.
Impacts causing plasticity are usually quite brief and they happen at distinct times. 
Simulation could allocate processing units for handling these and release them after impact has been processed.

>>>>

\subsubsection{Implementation using GPU}
\cbullet\ can already utilize GPU for many operations. 
Analysis on possible issues in used interfaces should be done before integrations.

@@

Current implementation is done based on \cbullet\ but it still needs refactoring and testing before it can be 
accepted to main stream. Interfaces should be reafctored so that handling of additional features like large 
displacements and material nonlinearity can be implemented without further changes.

@@

Update of constraints can be done in  many ways but using callbacks allows clean integration i.e. no changes
to solver core are needed.
In \cbullet\ there are already three per step callbacks.

@@

\subsection{Object subdivision just in time}
Just in time activation could avoid extra overhead from multiple objects and constraints.
Impacts causing plasticity are usually brief and they happen at distinct times. 
Simulation could allocate processing units for handling these and release them after impact has been processed.
